Parametricity for Primitive Nested Types and GADTs

### 2021-07-20 21-lmcs/21-lmcs.tex
MFPS revisions final(2)
just a test

$\mathit{pleaf}$,$\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms

@@

Considering $\relenv$ as a subcategory of the product $\Pi_{\phi^k \in
  \tvars \cup \fvars} RT_k$, computation of $\omega$-directed colimits
in $RT_k$ extends componentwise to $\relenv$. Recalling from the start
of this subsection that Definition~\ref{def:rel-sem} is given mutually
inductively with Definition~\ref{def:set-sem} we can now define our
relational interpretation.

\begin{defi}\label{def:rel-sem}
The {\em relational interpretation} $\relsem{\cdot} : \F \to [\relenv,
 \rel]$ is defined by
\begin{align*}
  \relsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \relsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \relsem{\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
  : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
    \Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
  &=
  \{(t,t') \in \setsem{\Gamma; \emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\emptyset
    \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
  \rho)~|~\\ 
  & \hspace{0.3in} \forall {R_1 : \rel(A_1,B_1)}\,...\,{R_k : \rel(A_k,B_k)}.\\
  & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
  (\relsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
      R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
  \relsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
  (\rho\phi)\ol{\relsem{\Gamma;\Phi \vdash 
    F}\rho}\\
  \relsem{\Gamma;\Phi \vdash F+G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho +
  \relsem{\Gamma;\Phi \vdash G}\rho\\
  \relsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho \times
  \relsem{\Gamma;\Phi \vdash G}\rho\\  
   \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho
  &= (----\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
  \text{where }	T_{H,\rho}
    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
  \text{and } T^\rel_{H,\rho}\,F
    &= \lambda \ol{R}. \relsem{
      \Gamma;\phi,\ol{\alpha} \vdash H}\rho[\phi :=
    F][\ol{\alpha := R}]\\
  \text{and } T^\rel_{H,\rho}\,\delta
    &= \lambda \ol{R}. \relsem{
      \Gamma;\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
    \delta][\ol{\alpha := \id_{\ol{R}}}]
\end{align*}
\end{defi}

@@

Theorem~\ref{eq:ft-from-nat} gives rise to an entire family of free
theorems that are consequences of naturality, and thus do not require
the full power of parametricity. In particular,
Definition~\ref{def:set-interp} alone ensures that standard properties
of the initial algebraic constructs $\map$, $\tin$, and $\fold$
hold. We have, for example, that the interpretation of every $\map_H$
is a functor, i.e., if $\Gamma; \ol\alpha, \ol\gamma \vdash H$,
$\ol{\Gamma; \emptyset \,|\, \Delta \vdash g : \Nat^{\ol\gamma} F \,
  G}$, and $\ol{\Gamma; \emptyset \,|\, \Delta \vdash f :
  \Nat^{\ol\gamma} G \, K}$, then
  \begin{align*}
&\setsem{
\Gamma; \emptyset \,|\, \Delta \vdash 
  (\map^{\ol{F}, \ol{K}}_H \,)_{\emptyset} \, \ol{(f \circ g)}
  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]} \\ 
= \hspace{0.03in} 
  &\setsem{
  \Gamma; \emptyset \,|\, \Delta \vdash 
  (\map^{\ol{G}, \ol{K}}_H \,)_{\emptyset} \, \ol{f} \circ 
  (\map^{\ol{F}, \ol{G}}_H \,)_{\emptyset} \, \ol{g}
  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]}
\end{align*}
In fact, we have that if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then,
for all $\ol{f \in \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset
    F\,G}\rho}$,
\[\begin{array}{rl}
&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
  \vdash (\map^{\ol{F},\ol{G}}_H)_\emptyset\,\ol x :
  \Nat^{\emptyset}\,H[\ol{\alpha := F}]\,H[\ol{\alpha := G}]}
\rho\,\ol f\\ =& \setsem{\Gamma; \ol{\alpha}\vdash H}
\id_{\rho}[\ol{\alpha := f}] \\ =&
\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
\end{array}\]
Here, we obtain the first equality from the appropriate instance of
Definition~\ref{def:set-interp}, and the second one by noting that
$\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
[\ol{\alpha := A}]$ is a functor in $\ol{\alpha}$ and using
$\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
$G$ on morphisms $\ol f$. We also have, e.g., that:

@@

Generalized algebraic datatypes (GADTs) --- also known as guarded
recursive data types~\cite{xcc03} or first-class phantom
types~\cite{ch03} --- generalize nested types by relaxing the above
restriction to allow the return types of data constructors to be
different instances of the datatype than the one being defined. For
example, the GADT
\[\begin{array}{l}
\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
\hspace*{0.4in}\mathtt{spair\;:\;Seq\;A \rightarrow Seq\;B \rightarrow
  Seq\;(A\times B)}\\
\hspace*{0.4in}\mathtt{sseq\;:\;(Nat \rightarrow Seq\;A) \rightarrow
  Seq\;(Nat \to A)}
\end{array}\]
\noindent
\!\!has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
return types $\mathtt{Seq\; (A \times B)}$ and $\mathtt{Seq\;(Nat \to
  A)}$. These types are not only at different instances of \verb|Seq|
from the instance $\mathtt{Seq\; A}$ being defined, but also at
different instances from one another. The resulting interdependence of
different instances of GADTs means that they can express more
constraints than ADTs and nested types. For example, the ADT
$\mathtt{List}$ expresses the invariant that all of the data in the
lists it defines is of the same type, while the nested type
$\mathtt{PTree}$ expresses this invariant {\em as well as} the
invariant that all of the lists it defines have lengths that are
powers of 2. The GADT $\mathtt{Seq}$ enforces {\em even more general}
well-formedness conditions for sequences of values that simply cannot
be expressed with ADTs and nested types alone.

@@

\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
      K}\rho {[\ol{\alpha := A} ]}}\;\;\;\;"{left}] && \set \\  
&\set^{|\ol K|} \ar[ur, "\;\;\mathit{Lan}_{\ol{\lambda \ol A.\,
      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
\end{tikzcd}\]
and, for all $\ol{N : RT_k}$ and $d : \setsem{\Gamma;\emptyset \vdash
  \Delta}$,
\[\begin{array}{l}
\mu_{\ol{N}\, d} : 
\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
\hspace*{0.8in} \to \lambda \ol B.\, \setsem{\Gamma; 
  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
\end{array}\]
is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
\end{equation}
as given by the universal property of the left Kan extension.  The
relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
  G}$ are defined entirely analogously.

@@

\noindent
{\small \sc Acknowledgment} We thank Daniel Jeffries for his
collaboration on~\cite{jgj21}, for his contributions to a preliminary
version of this paper, and for supporting publication of the present
version in his absence.

>>>>

$\mathit{pleaf}$, $\mathit{pnode}$, $\mathit{swap}$, and $s$ are the terms

@@

Considering $\relenv$ as a subcategory of the product $\Pi_{\phi^k \in
  \tvars \cup \fvars} RT_k$, computation of $\omega$-directed colimits
in $RT_k$ extends componentwise to $\relenv$. Recalling from the start
of this subsection that Definition~\ref{def:rel-sem} is given mutually
inductively with Definition~\ref{def:set-sem} we can now define our
relational interpretation. As with the set interpretation, the
relational interpretation is given in two parts, in
Definitions~\ref{def:rel-sem} and~\ref{def:rel-sem-funcs}.

\begin{defi}\label{def:rel-sem}
The {\em relational interpretation} $\relsem{\cdot} : \F \to [\relenv,
 \rel]$ is defined by
\begin{align*}
  \relsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \relsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \relsem{\Gamma; \emptyset \vdash \Nat^{\ol{\alpha}} \,F\,G}\rho &= \{\eta
  : \lambda \ol{R}.\,\relsem{\Gamma; \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := R}] \Rightarrow \lambda \ol{R}. \,\relsem{
    \Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha := R}]\}\\
  &=
  \{(t,t') \in \setsem{\Gamma; \emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\emptyset
    \vdash \Nat^{\ol{\alpha}} \,F\,G} (\pi_2
  \rho)~|~\\ 
  & \hspace{0.3in} \forall {R_1 : \rel(A_1,B_1)}\,...\,{R_k : \rel(A_k,B_k)}.\\
  & \hspace{0.4in} (t_{\ol{A}},t'_{\ol{B}}) \in
  (\relsem{\Gamma; \ol{\alpha} \vdash G}\rho[\ol{\alpha :=
      R}])^{\relsem{\Gamma;\ol{\alpha}\vdash F}\rho[\ol{\alpha := R}]} \}\\  
  \relsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
  (\rho\phi)\ol{\relsem{\Gamma;\Phi \vdash 
    F}\rho}\\
  \relsem{\Gamma;\Phi \vdash F+G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho +
  \relsem{\Gamma;\Phi \vdash G}\rho\\
  \relsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \relsem{\Gamma;\Phi \vdash F}\rho \times
  \relsem{\Gamma;\Phi \vdash G}\rho\\  
   \relsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho
  &= (\mu T_{H,\rho})\ol{\relsem{\Gamma;\Phi \vdash G}\rho}\\
  \text{where }	T_{H,\rho}
    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
  \text{and } T^\rel_{H,\rho}\,F
    &= \lambda \ol{R}. \relsem{
      \Gamma;\phi,\ol{\alpha} \vdash H}\rho[\phi :=
    F][\ol{\alpha := R}]\\
  \text{and } T^\rel_{H,\rho}\,\delta
    &= \lambda \ol{R}. \relsem{
      \Gamma;\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
    \delta][\ol{\alpha := \id_{\ol{R}}}]
\end{align*}
\end{defi}

@@

Theorem~\ref{eq:ft-from-nat} gives rise to an entire family of free
theorems that are consequences of naturality, and thus do not require
the full power of parametricity. In particular,
Definition~\ref{def:set-interp} alone ensures that standard properties
of the initial algebraic constructs $\map$, $\tin$, and $\fold$
hold. We have, for example, that the interpretation of every $\map_H$
is a functor, i.e., if $\Gamma; \ol\alpha, \ol\gamma \vdash H$,
$\ol{\Gamma; \emptyset \,|\, \Delta \vdash g : \Nat^{\ol\gamma} F \,
  G}$, and $\ol{\Gamma; \emptyset \,|\, \Delta \vdash f :
  \Nat^{\ol\gamma} G \, K}$, then
  \begin{align*}
&\setsem{
\Gamma; \emptyset \,|\, \Delta \vdash 
  (\map^{\ol{F}, \ol{K}}_H \,)_{\emptyset} \, \ol{(f \circ g)}
  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]} \\ 
= \hspace{0.03in} 
  &\setsem{
  \Gamma; \emptyset \,|\, \Delta \vdash 
  (\map^{\ol{G}, \ol{K}}_H \,)_{\emptyset} \, \ol{f} \circ 
  (\map^{\ol{F}, \ol{G}}_H \,)_{\emptyset} \, \ol{g}
  : \Nat^{\ol\gamma} H[\ol{\alpha := F}] \, H[\ol{\alpha := K}]}
\end{align*}
In fact, we have that if $\Gamma; \ol{\alpha} \vdash H$, $\ol{\Gamma;
  \emptyset \vdash F}$, and $\ol{\Gamma; \emptyset \vdash G}$, then,
for all $\ol{f \in \setsem{\Gamma;\emptyset \vdash \Nat^\emptyset
    F\,G}\rho}$,
\[\begin{array}{rl}
&\setsem{ \Gamma;\emptyset\,|\,\ol{x : \Nat^\emptyset F \, G}
  \vdash (\map^{\ol{F},\ol{G}}_H)_\emptyset\,\ol x :
  \Nat^{\emptyset}\,H[\ol{\alpha := F}]\,H[\ol{\alpha := G}]}
\rho\,\ol f\\ =& \setsem{\Gamma; \ol{\alpha}\vdash H}
\id_{\rho}[\ol{\alpha := f}] \\ =&
\mathit{map}_{\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho [\ol{\alpha := A}]} \ol{f} \\
\end{array}\]
Here, we obtain the first equality from the appropriate instance of
Definition~\ref{def:set-interp}, and the second one by noting that
$\lambda \ol{A}.\,\setsem{\Gamma; \ol{\alpha}\vdash H} \rho
[\ol{\alpha := A}]$ is a functor in $\ol{A}$ and using
$\mathit{map}_G\,\ol f$ to denote the action of the semantic functor
$G$ on morphisms $\ol f$. We also have, e.g., that:

@@

Generalized algebraic data types (GADTs) --- also known as guarded
recursive data types~\cite{xcc03} or first-class phantom
types~\cite{ch03} --- generalize nested types by relaxing the above
restriction to allow the return types of data constructors to be
different instances of the data type than the one being defined. For
example, the GADT
\[\begin{array}{l}
\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
\hspace*{0.4in}\mathtt{spair\;:\;Seq\;A \rightarrow Seq\;B \rightarrow
  Seq\;(A\times B)}\\
\hspace*{0.4in}\mathtt{sseq\;:\;(Nat \rightarrow Seq\;A) \rightarrow
  Seq\;(Nat \to A)}
\end{array}\]
\noindent
has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
return types $\mathtt{Seq\; (A \times B)}$ and $\mathtt{Seq\;(Nat \to
  A)}$. These types are not only at different instances of \verb|Seq|
from the instance $\mathtt{Seq\; A}$ being defined, but also at
different instances from one another. The resulting interdependence of
different instances of GADTs means that they can express more
constraints than ADTs and nested types. For example, the ADT
$\mathtt{List}$ expresses the invariant that all of the data in the
lists it defines is of the same type, while the nested type
$\mathtt{PTree}$ expresses this invariant {\em as well as} the
invariant that all of the lists it defines have lengths that are
powers of 2. The GADT $\mathtt{Seq}$ enforces {\em even more general}
well-formedness conditions for sequences of values that simply cannot
be expressed with ADTs and nested types alone.

@@

\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
      K}\rho {[\ol{\alpha := A} ]}}\;\;\;\;"{left}] && \set \\  
&\set^{|\ol K|} \ar[ur, "\;\;\mathit{Lan}_{\ol{\lambda \ol A.\,
      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
\end{tikzcd}\]
and, for all $\ol{N : RT_k}$ and $d : \setsem{\Gamma;\emptyset \vdash
  \Delta}$,
\[\begin{array}{l}
\mu_{\ol{N}\, d} : 
\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
\hspace*{0.8in} \to \lambda \ol B.\, \setsem{\Gamma; 
  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
\end{array}\]
is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
\end{equation}
as given by the universal property of the left Kan extension.  The
relational interpretations\\ $\relsem{\Gamma;\emptyset~|~\emptyset
  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
  G}$ are defined entirely analogously.

@@

\noindent
{\small \sc Acknowledgment} We thank Daniel Jeffries for his
collaboration on~\cite{jgj21}, for his contributions to a preliminary
version of this paper, and for supporting publication of the present
version in his absence. We also thank the anonymous reviewers for
their detailed and thorough reviews, which have greatly improved the
paper.
